% Synthos: A Self-Evolving Cognitive Operating System for AI-Augmented Scientific Research
\documentclass[review]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{textcomp}
\usepackage[utf8]{inputenc}
\usepackage{array}
\usepackage{algorithm}
\usepackage{algpseudocode}

\journal{Intelligent Computing}

\begin{document}

\begin{frontmatter}

\title{Synthos: A Self-Evolving Cognitive Operating System for AI-Augmented Scientific Research}

\author[1]{Xiaokai Yang\textsuperscript{*}}
\ead{yakeworld@outlook.com}
\author[2]{Synthos Project}
\address[1]{Department of Neurology, Wenzhou People's Hospital, Wenzhou 325000, China}
\address[2]{GitHub: \url{https://github.com/yakeworld/Synthos}}

\begin{abstract}
Large language model (LLM) agents have demonstrated remarkable potential in scientific research assistance, yet existing systems remain fragmented between tool orchestration and genuine cognitive reasoning. We present Synthos, a self-evolving cognitive operating system that reifies eight epistemological principles as executable SKILL.md cognitive atoms, governed by a formal constitutional hierarchy (CON~$\gg$~MEM~$\gg$~CMD~$\gg$~SKL~$\gg$~DEF). The system's architecture is formally defined through a cognitive atom as a typed I/O function, an evolution engine as a finite state machine with five conditional gates and three execution paths, and an absorption pipeline as a multi-stage filter with utility-threshold gating. Over 48 evolution cycles, Synthos achieved a composite quality score of 0.98 (EXCELLENT), maintained all cognitive atoms at perfect structural and benchmark scores of 1.00, absorbed patterns from 9 external projects, and produced 40+ scientific paper drafts spanning ophthalmology, neurology, biomechanics, and AI. We provide formal definitions of the evolution state machine, the constitutional priority algebra, and the cognitive atom I/O contracts, establishing a rigorous foundation for self-evolving scientific AI systems.
\end{abstract}

\begin{keyword}
Cognitive Architecture \sep Self-Evolution \sep Scientific AI Agent \sep Constitutional AI \sep Epistemological Computing
\end{keyword}
\end{frontmatter}

\section{Introduction}
\label{sec:introduction}

\subsection{The Rise of AI Research Agents}
\label{sec:rise}

The pursuit of autonomous scientific discovery through artificial intelligence has produced a diverse ecosystem of research agents~\cite{elovic2024gpt,sakana2025ai,shao2024storm,futurehouse2025paperqa2,asai2025openscholar}. These systems have achieved remarkable benchmarks---PaperQA achieving 86.3\% on PubMedQA, AI Scientist producing the first AI-written workshop paper---yet they share a fundamental limitation: their cognitive architecture remains implicit, embedded in orchestration code rather than explicitly modeled as a reasoning framework.

\subsection{Three Critical Gaps}
\label{sec:gaps}

\textbf{Epistemological rigor is weakly enforced.} Most agents lack formal mechanisms for evidence traceability, falsifiability constraints, or systematic bias detection. The SKILL.md paradigm~\cite{hermes2025hermes} introduced executable skill files, but its application to scientific reasoning remains nascent. While cognitive architectures for language agents have been surveyed~\cite{sumers2023cognitive} and generative agent frameworks proposed~\cite{park2023generative}, comprehensive agent surveys~\cite{wang2024surveyagent,xi2023rise} confirm that none formally encode epistemological constraints. Critically, even systems that do invoke epistemological concepts---such as Claude Code's constitutional drift detection~\cite{anthropic2025claude}---restrict their scope to narrow domains (code generation), never elevating epistemological constraints to first-class architectural components governing the full research lifecycle. Notably, traditional symbolic cognitive architectures---SOAR~\cite{laird2012soar} with its rule-based production system and ACT-R~\cite{anderson2004actr} with its modular buffer architecture---established the principle that cognition can be decomposed into structured operations with explicit control flow, yet their fixed-rule representations lack the semantic flexibility to adapt to novel scientific domains without extensive manual reprogramming. Synthos inherits this symbolic tradition while replacing rigid production rules with LLM-interpretable declarative SKILL.md atoms.

\textbf{Self-improvement remains ad-hoc, and imperatively locked.} Evolution in existing systems is typically limited to tool-patch updates rather than architectural self-reflection. LangGraph~\cite{langchain2024langgraph} introduced state machines for agent workflows, LangChain~\cite{chase2022langchain} pioneered composable LLM applications, and DSPy~\cite{khattab2023dspy} demonstrated automated prompt optimization. However, these frameworks share a fundamental architectural limitation: all control flow---task decomposition, conditional branching, iteration---is hard-coded in Python\@. The LLM functions as a ``called subroutine'' within an imperative harness, incapable of reading, understanding, or modifying the orchestration logic that governs its own reasoning. A MetaGPT~\cite{hong2023metagpt} or CrewAI~\cite{crewai2024crewai} agent cannot, at runtime, inspect its SOP definition and decide to restructure it, because the SOP is embedded in Python conditionals invisible to the LLM. Similarly, Auto-GPT~\cite{significantgravitas2023autogpt} and AI Scientist~\cite{sakana2025ai} achieve remarkable autonomy yet cannot apply falsificationist constraints to their own hypothesis generation---the LLM generates ideas within an imperative loop that has no epistemic gate. This is the \emph{imperative evolution lock-in}: as long as an agent's cognitive architecture is compiled into imperative code, the LLM is permanently excluded from reflecting on and restructuring its own reasoning framework. Techniques such as Voyager~\cite{wang2023voyager} and Reflexion~\cite{shinn2023reflexion} advance self-improvement via iterative prompting, but their corrective mechanisms remain confined within the Python orchestration layer. ORCA~\cite{mukherjee2023orca} demonstrates progressive learning through explanation chains but lacks any architecture-modifying capability. DSPy~\cite{khattab2023dspy} optimizes prompts automatically, yet the optimizer itself is a Python function.

\textbf{Philosophical principles remain aspirational.} Even systems that invoke the ``scientific method'' do not execute epistemological constraints as first-class architectural components. Foundational LLM advances---Transformers~\cite{vaswani2017attention}, GPT-3~\cite{brown2020language}, chain-of-thought reasoning~\cite{wei2022chain}, self-consistency~\cite{wang2023selfconsistency}, tree-of-thoughts~\cite{yao2023tree}, and retrieval-augmented generation~\cite{lewis2020retrieval}---have enabled increasingly capable agents, yet the reasoning frameworks that govern these capabilities remain implicit and non-enforceable.

\subsection{Related Work and Positioning}
\label{sec:related}

Recent advances address these gaps from complementary but incomplete directions. Claude Code~\cite{anthropic2025claude} introduced constitutional AI with drift detection, restricting its scope to code generation. DeepResearchAgent~\cite{skyworkai2026deep} proposed SEPL rollback protocols for self-evolution but remains Python-dependent. AI-Research-SKILLs~\cite{orchestra2026airesearch} demonstrated 90+ SKILL.md files with ARA provenance, yet lacks a self-evolution mechanism. GEPA~\cite{gepa2025gepa} introduced reflective strategy selection without constitutional governance. Most recently, Microsoft SkillOpt~\cite{yang2026skillopt} proposed a systematic text-space optimizer that treats skill documents as the external trainable state of a frozen agent---applying training-style controls (epochs, batch size, learning rate budgets, validation gates) to single-skill editing. SkillOpt achieves best-or-tied on all 52 evaluated (model, benchmark, harness) cells, surpassing GEPA and five other baselines, yet optimizes one skill at a time via an additional frontier model and targets structured benchmarks rather than open-ended scientific research. STORM~\cite{shao2024storm} excels at literature synthesis but does not extend to hypothesis generation or verification. OpenScholar~\cite{asai2025openscholar} synthesizes literature at scale but operates as a monolithic pipeline. Nature-skills~\cite{yuan2025nature} and Academic Research Skills~\cite{ars2025academic} advanced the SKILL.md paradigm for task-specific capabilities, while PaperOrchestra~\cite{song2026paperorchestra} demonstrated multi-agent pipeline advantages. AutoR~\cite{autox2025autor} shares the human-in-loop philosophy but implements it via Python harness. KILO-KIT~\cite{vo2025kilo} contributed CRISP-DM templates for structured research workflows.

The pattern across all surveyed systems reveals a common architectural bottleneck: imperative control flow. Whether implemented as Python orchestration (LangChain, CrewAI, AutoR), fixed SOP definitions (MetaGPT), or harnessed loops (AI Scientist), every existing system embeds its cognitive logic in a layer that the LLM cannot read, reflect upon, or restructure. The corrective mechanisms of Voyager and Reflexion remain trapped within this imperative harness---they can adjust prompts but cannot modify the orchestration architecture itself. Consequently, even the most autonomous agents are locked into their initial architectural paradigm, unable to evolve their reasoning framework beyond what a human developer hard-codes. This observation motivates a fundamental design shift: to enable genuine cognitive self-evolution, the system's entire architecture must be expressed in a format that is both executable by the LLM and readable/modifiable by the LLM---a declarative epistemological medium. Critically, complex epistemological constraints---such as Falsificationism's mandatory counterargument generation or Bayesian confidence propagation---would themselves become unwieldy and error-prone if hard-coded as imperative logic (nested conditions, callback chains, state variables); only when reified as natural-language declarative rules (CONSTITUTION.md priorities, SKILL.md cognitive atoms) can the LLM faithfully interpret and apply them through its native next-token-prediction mechanism, unifying the two design requirements---architectural self-modifiability and epistemic rigor---into a single coherent solution. No existing system simultaneously achieves explicit epistemological encoding, constitutional governance, self-evolving architecture, and zero-Python execution.

\subsection{Our Contributions}
\label{sec:contributions}

\textbf{Contribution 1: Epistemological Code} --- Eight philosophical frameworks from the history and philosophy of science are reified as executable SKILL.md cognitive atoms with formal I/O contracts, building on representation learning principles~\cite{bengio2013representation}. First Principles governs decomposition; Falsificationism ensures every hypothesis generates explicit counterargument paths; Bayesian Reasoning drives confidence propagation across the verification chain.

\textbf{Contribution 2: Constitutional Hierarchy} --- All system behavior is governed by CONSTITUTION.md v5.0~\cite{yang2026synthosconst}, a formal hierarchy of immutable principles ranked CON~$>$~MEM~$>$~CMD~$>$~SKL~$>$~DEF, enforced by a philosophical immune system with formal conflict resolution.

\textbf{Contribution 3: Self-Evolution Infrastructure} --- An 11-step evolution engine~\cite{yang2026synthos} (v2.11) with formal state machine definition, incorporating PW-Bench~\cite{song2026paperorchestra}, SEPL rollback~\cite{skyworkai2026deep}, Git-as-Memory~\cite{haiyang2025aris}, hypothesis-first single-metric improvement, and active inference gates.

Table~\ref{tab:comparison} compares Synthos with representative existing systems across key dimensions.

\begin{table}[h]\centering\caption{Comparison with representative research agent systems.}\label{tab:comparison}
\small
\begin{tabular}{lcccc}\toprule
\textbf{System} & \textbf{Epistemic} & \textbf{Const.} & \textbf{Self-Evol.} & \textbf{Zero-Py}\\\midrule
GPT-Researcher~\cite{elovic2024gpt} & $\times$ & $\times$ & $\times$ & $\times$\\
AI Scientist~\cite{sakana2025ai} & $\times$ & $\times$ & $\bigtriangleup$ & $\times$\\
PaperQA2~\cite{futurehouse2025paperqa2} & $\times$ & $\times$ & $\times$ & $\times$\\
Claude Code~\cite{anthropic2025claude} & $\times$ & $\checkmark$ & $\times$ & $\times$\\ 
DSPy~\cite{khattab2023dspy} & $\times$ & $\times$ & $\checkmark$ & $\times$\\
DeepResearchAgent~\cite{skyworkai2026deep} & $\times$ & $\times$ & $\checkmark$ & $\times$\\
AI-Research-SKILLs~\cite{orchestra2026airesearch} & $\bigtriangleup$ & $\times$ & $\times$ & $\checkmark$\\
Voyager~\cite{wang2023voyager} & $\times$ & $\times$ & $\bigtriangleup$ & $\times$\\
ORCA~\cite{mukherjee2023orca} & $\times$ & $\times$ & $\bigtriangleup$ & $\times$\\
SkillOpt~\cite{yang2026skillopt} & $\times$ & $\times$ & $\checkmark$ & $\checkmark$\\
\textbf{Synthos (ours)} & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\\bottomrule
\end{tabular}
\end{table}

\section{Architecture}
\label{sec:architecture}

\subsection{Design Principles}

Synthos is built on four immutable principles~\cite{yang2026synthosconst}: (1) Carbon-Silicon Symbiosis---human-in-the-loop; (2) Self-Evolution as First Principle---all outputs serve evolution; (3) Precise Atomic Boundaries---non-overlapping cognitive atoms; (4) Epistemological Honesty Over Sycophancy---claims cite sources. These form a priority hierarchy: CON~$\gg$~MEM~$\gg$~CMD~$\gg$~SKL~$\gg$~DEF, enforced by a philosophical immune system (distinct from the RLAIF weight-update paradigm of original Constitutional AI~\cite{bai2022constitutional}---Synthos enforces its constitution via deterministic runtime parsing of SKILL.md directives, not RL-based model training).

\end{document}
